---
title: "Using R to analyse a simple survey"
author: "Nathaniel Phillips"
date: "24 March 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## The survey

For these examples, we'll load the following data representing psychology study where 150 participants were assigned to one of 3 conditions A, B, and C. For each participant, their response to some measure was recorded, along with their accuracy, and the time of their response. The data are contained in a text file called `study.txt` at [https://raw.githubusercontent.com/ndphillips/PsyKo-2017/master/data/study.txt](https://raw.githubusercontent.com/ndphillips/PsyKo-2017/master/data/study.txt)


Here are the first few rows of the data:

```{r, echo = FALSE, eval = TRUE}
# Set working directory to directory containing the .Rproj file
setwd(rprojroot::is_rstudio_project$find_file())

# Example: Loading data from a text file called mydata.txt into a new object called x

study <- read.table(file = "data/study.txt",      # The file is called mydata.txt and is in the working directory
                 sep = "\t",               # Columns in the data are separated by tabs
                 header = TRUE,            # There is a header row
                 stringsAsFactors = FALSE) # Do not convert strings to Factors

knitr::kable(head(study))
```


## Installing packages

To use functions or data from a package, like `yarrr` or `BayesFactor`, you first must install the package from the internet on your computer with `install.packages()`. Note, you must be connected to the internet to install the package, but you only need to do this once.

```{r}
# Install some packages from the internet (you only need to install packages to your computer once)

install.packages("yarrr")         # For R piratery
install.packages("BayesFactor")   # For Bayesian analyses
install.packages("rprojroot")     # For working directory handling
install.packages("tidyverse")     # For data wrangling
install.packages("devtools")      # For accessing github
install.packages("knitr")         # For document creation
install.packages("rmarkdown")     # For markdown documents
devtools::install_github("crsh/papaja") # APA analyses
install.packages("tm")                  # Natural language processing
```


## Loading packages

To use a package you've installed in your code, you must always load it with `library()`

```{r}
# Load packages for my analyses. You must always load all packages you plan to use!

library("yarrr")         # For R piratery
library("BayesFactor")   # For Bayesian analyses
library("papaja")        # For APA style results
library("rprojroot")     # For working directory handling
library("tidyverse")     # For data wrangling
library("papaja")        # APA analyses
library("tm")            # natural language processing
```



## Loading data


Now let's load the study data as a new object called `study`. The data are stored as a text file called `study.txt`. To load the data into R, we'll use the `read.table()` function.

| Argument| Description| 
|:------------|:-------------------------------------------------|
|`file`| The document's file path relative to the working directory unless specified otherwise. For example `file = "mydata.txt"` looks for the text file directly in the working directory, while `file = "data/mydata.txt"` will look for the file in an existing folder called `data` inside the working directory.<br>If the file is outside of your working directory, you can also specify a full file path (`file = "/Users/CaptainJack/Desktop/OctoberStudy/mydata.txt"`) |
|`header`|  A logical value indicating whether the data has a header row -- that is, whether the first row of the data represents the column names.|
|`sep`|  A string indicating how the columns are separated. For comma separated files, use `sep = ","`, for tab--delimited files, use `sep = "\t"` |
|`stringsAsFactors`|  A logical value indicating whether or not to convert strings to factors. I **always** set this to FALSE because I *hate*, *hate*, *hate* how R uses factors|


```{r, eval = TRUE}
# Set working directory to directory containing an .Rproj file
setwd(rprojroot::is_rstudio_project$find_file())

# Example: Loading data from a text file called mydata.txt into a new object called x

study <- read.table(file = "data/study.txt", # The file is called mydata.txt and is in the working directory
                 sep = "\t",                 # Columns in the data are separated by tabs
                 header = TRUE,              # There is a header row
                 stringsAsFactors = FALSE)   # Do not convert strings to Factors

# Note, you can also read it directly from the internet as follows:
study <- read.table(file = "https://raw.githubusercontent.com/ndphillips/PsyKo-2017/master/data/study.txt",
                 sep = "\t",                 # Columns in the data are separated by tabs
                 header = TRUE,              # There is a header row
                 stringsAsFactors = FALSE)   # Do not convert strings to Factors
```



## Exploring data 

When you first load a dataset, you should take a look at it to make sure it was loaded correctly. Here are common functions for getting to know a new dataset:

| Function| Description| 
|:------------------------|:-----------------------------|
|     `head(study), tail(study)`| Print the first few rows (or last few rows). | 
|     `View(study)`| Open the entire object in a new window | 
|     `nrow(study), ncol(study), dim(study)`| Count the number of rows and columns  | 
|     `rownames(), colnames(), names()`| Show the row (or column) names   | 
|     `str(study), summary(study)`| Show the structure of the dataframe (ie., dimensions and classes) and summary statistics| 



```{r, eval = FALSE}
head(study)   # Show me the first few rows of study
View(study)   # View the entire dataframe in a new window
str(study)    # Show me basic information about the dataframe
dim(study)    # How many rows and columsn are there?
```


## Working with column names
To access columns in a dataframe by name, use the `$` operator. For example, `study$sex` will return the sex values in a dataframe called study

```{r, eval = FALSE}
names(study)                    # Show me the names of the columns in study
study$age                       # Return the column age in the dataframe
study$age.months <- study$age * 12  # Add a new column age.months to the dataframe
names(study)[8] <- "months"    # Change the name of the 7th column to "months"
```


## Descriptive statistics functions

### Continuous, numeric data

These functions will return descriptive statistics for numeric data like `age` and `time`:

```{r}
sum(study$age)
min(study$response)
max(study$time)
mean(study$time)
median(study$response)
sd(study$time)
var(study$age)
summary(study$time)
```

### Categorical data

These functions will return descriptive statistics for categorical data like `sex` and `condition`:

```{r}
unique(study$age)                       # What are all unique values of age?
table(study$condition, exclude = NULL)  # How many times did each condition value occur
```

### Missing (NA) Values

If your data contains missing (NA) values, you may need to include additional arguments in your functions to prevent errors:

```{r}
# include na.rm = TRUE to ignore missing values

mean(study$response, na.rm = TRUE)   # Show me the mean of response, but ignore missing (NA) values
max(study$time, na.rm = TRUE)        # Show me the median of time, but ignore NA values

table(study$sex, exclude = NULL)     # Show me a table of the values of sex, and include counts of NA values
```

## Grouped aggregation

To calculate multiple descriptive statistics separated by groups of data, you can use the `aggregate()` function:

```{r}
# Show me the mean age for each sex
aggregate(formula = age ~ sex, 
          data = study, 
          FUN = mean)


# Show me the median response for each condition, but only for males
aggregate(formula = response ~ condition,
          data = study,
          subset = sex == "m",
          FUN = median)


# Show me the mean time for each condition and sex,
aggregate(formula = time ~ condition + sex,
          data = study,
          FUN = mean)
```

Alternatively, you can use the `dplyr` package to do more advanced data aggregation:

```{r}
# Give me several summary statistics for each combination of condition and sex:

study %>% group_by(condition) %>%
  summarise(
    N = n(),
    age.mean = mean(age),
    percent.male = mean(sex == "m"),
    response.mean = mean(response),
    response.median = median(response),
    accuracy.mean = mean(accuracy)
    )
```


## Plotting

### Colors

R has many ways of defining colors, here are some of them:

```{r}
colors()                                      # See all of the named colors in R
demo("colors")                                # Run a demo of colors
yarrr::piratepal()                            # Show the yarrr color palettes
yarrr::piratepal("pony", plot.result = TRUE)  # Show the pony palette
```

### Histograms

Histograms give you a quick view of the distribution of a vector of numeric data:
```{r}
# Histogram of responses
hist(study$response)

# Fancier histogram
hist(study$response, 
     xlim = c(0, 100),               # x axis limits
     breaks = 20,                    # number of bins
     col = "dodgerblue",             # Color of bins
     border = "white",               # Bin border color
     main = "Response Distribution",
     xlab = "Response", 
     yaxt = "n",                     # turn off y-axis
     ylab = "")
```


### Scatterplots

Scatterplots show you the relationship between two continuous, numeric variables:
```{r}
# Scatterplot of responses and time
plot(x = study$response, 
     y = study$time)


# Fancier scatterplot of reponse and time
plot(x = study$response, 
     y = study$time, 
     xlab = "Response",
     ylab = "Time (milliseconds)",
     pch = 16,                      # Solid points
     col = gray(.5, .5),            # Transparent gray points
     xlim = c(0, 100))

grid()                              # Add gridlines

# Add a regression line
abline(lm(time ~ response, 
          data = study),
       lty = 2)                    # Dashed line
```

### Barplot

Barplots show you the relationship between a categorical independent variable (like condition or sex), and a continuous dependent variable (like response):

```{r}
# Barplot of response by condition using papaja::apa_barplot
papaja::apa_barplot(data = study, 
                    id = "id", 
                    factors = c("condition"), 
                    dv = "accuracy")

# Now including two IVs
papaja::apa_barplot(data = study, 
                    id = "id",                           # Column indicating individual participants
                    factors = c("condition", "sex"),     # What are the IVs?
                    dv = "accuracy",                     # What is the DV
                    args_legend = list(x = "topleft"))   # Put legend on top left
```


### pirateplots

Pirateplots are advanced versions of barplots that allow you to easy visualize an entire distribution of data:

```{r}
# Pirateplot of response by condition
yarrr::pirateplot(response ~ condition,
                  data = study)

#  Theme 2
yarrr::pirateplot(response ~ condition,
                  data = study, 
                  theme = 2, 
                  back.col = "oldlace")

# Theme 3 and 2 IVs
yarrr::pirateplot(response ~ condition + sex,
                  data = study, 
                  pal = "google",              # use the google color palette
                  theme = 3)
```


## Hypothesis tests

### t-tests

```{r}
# One sample t-test comparing the mean response to 100
t.test(study$response, 
       alternative = "two.sided",
       mu = 500)

# Two sample t-test comparing the mean response of males to females
t.test(response ~ sex,
       data = study,
       alternative = "two.sided")

# Two sample t-test comparing the mean time between condition A and B
t.test(time ~ condition,
       data = study,
       subset = condition %in% c("A", "B"),
       alternative = "two.sided")
```

### Correlation test

```{r}
# Correlation test between age and time
cor.test(formula = ~ age + time,
         data = study)

# Correlation test between response and time, but only for females
cor.test(formula = ~ response + time,
         data = study,
         subset = sex == "f")
```

### ANOVA

```{r}
# ANOVA comparing response by condition
response.aov <- aov(formula = response ~ condition,
                    data = study)

# Summary results
summary(response.aov)

# Post-hoc tests
TukeyHSD(response.aov)
```


### Regression

```{r}
# Regression analysis predicting response by age, sex and condition
response.lm <- lm(formula = response ~ age + sex + condition,
                  data = study)

# Summary results
summary(response.lm)
```


## Bayesian Statistics

Bayesian statistics provide a much richer interpretation of your data than standard non-Bayesian 'frequentist' statistics. In R, you can easily perform Bayesian versions of your favorite statistical analyses (like t-tests, ANOVAs, and regression) using the `BayesFactor` package. For full details on the `BayesFactor` package, look at the great package manual by running `BFManual()`

```{r}
# Bayesian t-test comparing times of men and women
time.bf <- BayesFactor::ttestBF(formula = time ~ sex, 
                                data = study)

time.bf       # Show results
plot(time.bf)

# Bayesian ANOVA comparing reponse by condition

study$condition.f <- factor(study$condition)  # Convert condition to a factor

response.bf <- BayesFactor::anovaBF(formula = response ~ condition.f, 
                                    data = study)

response.bf       # Show results
plot(response.bf)

# Bayesian ANOVA comparing reponse by condition and age
study$age.f <- factor(study$age)           # Convert condition to a factor

response.B.bf <- BayesFactor::anovaBF(formula = response ~ condition.f + age.f, 
                                      data = study)

response.B.bf       # Show results
plot(response.B.bf)
```


### Simulations


#### Simulation 1: What is the distribution of p-values given a specified mean and sd?

- If we take a sample size of size N from a normal distribution with mean `true.mean` and standard deviation `true.sd`, and conduct a one-sample t-test, what will the distribution of p-values be?

```{r, eval = FALSE}
# SIMULATION 1
#   What is the distribution of p-values given a specified mean and sd?

# --------------------
# Determine simulation parameters
# --------------------

true.mean <- 0     # The true population mean
true.sd <- 1       # The true population standard deviation
N <- 10           # The number of samples in each simulation
sim <- 1000        # Number of simulations

# --------------------
# Start pulation!
# --------------------

# Create a place-holder vector for the p-values
p <- rep(NA, sim)

# Loop over simulations
for(i in 1:sim) {
  
  # Generate random data
  data <- rnorm(n = N, mean = true.mean, sd = true.sd)
  
  # Calculate a one-sample t-test
  test <- t.test(data)
  
  # Get the p-value
  p.value <- test$p.value
  
  # Assign the p-value to the ith place in p
  p[i] <- p.value
  
}

# --------------------
# Show the results!
# --------------------

# What percent are less than or equal to .05?
mean(p < .05)

# Histogram of distribution of p-values
hist(p, 
     xlim = c(0, 1), 
     xlab = "p-value", 
     main = paste("p(p-value < .05) =", round(mean(p < .050), 2)), 
     probability = TRUE, yaxt = "n", ylab = "")

# Add raw data

points(x = p, 
       y = rep(.5, length(p)) + rnorm(length(p), mean = 0, sd = .05),
       pch = 16,
       col = gray(0, .1), cex = 1.5)

abline(v = .05, lty = 2)
```

#### Simulation 2: Given that a p-value is 'significant' what is the probability that the null hypothesis is actually false?

- If a p-value is significant, what is the probability that the null hypothesis is actually false?
- To answer this, we'll draw random samples *either* from a distribution where the null hypothesis is true, or where the null hypothesis is false. 
- We'll then conduct a t-test and see if the result is significant.
- After we conduct many simulations, we'll calculate the pr

```{r}
# SIMULATION 2
# Given that a p-value is 'significant' what is the probability that 
#  the null hypothesis is actually false?

# --------------------
# Determine simulation parameters
# --------------------

n.H0 <- 500      # The number of simulations where the null is TRUE
n.H1 <- 500      # The number of simulations where the null is FALSE

N <- 100         # The sample size for each simulation
mean.H0 <- 0     # The mean when the null is TRUE (must be 0!)
mean.H1 <- .25   # The mean when the null is FALSE
sd.H0.H1 <- 1    # The standard deviation of all distributions

# --------------------
# Start simulation!
# --------------------

total.simulations <- n.H0 + n.H1

# Create a dataframe of parameters and a placeholder for the p-values
sim <- data.frame(mu = c(rep(mean.H0, n.H0), rep(mean.H1, n.H1)),
                  sd = rep(sd.H0.H1, n.H0 + n.H1),
                  p = NA)

# Loop over all simulations
for(i in 1:total.simulations) {
  
  # Get the true mean and standard deviation for the ith simulation
  true.mu <- sim$mu[i]
  true.sd <- sim$sd[i]
  
  # Generate random data
  data <- rnorm(n = N, mean = true.mu, sd = true.sd)
  
  # Conduct a one-sample t-test
  test <- t.test(data)
  
  # Get the p-value
  p.value <- test$p.value
  
  # Assign the p-value to the ith place in the dataframe
  sim$p[i] <- p.value
  
}

# --------------------
# Show the results!
# --------------------

# If the null is TRUE, what is the probability of a significant result?
mean(sim$p[sim$mu == 0] < .05)

# If the null is FALSE, what is the probability of a significant result?
mean(sim$p[sim$mu != 0] < .05)

# Given a significant p-value, what is the probability that the null hypothesis is false?
with(subset(sim, p < .05), mean(mu != 0))

## Plotting

par(mar = c(5, 6, 4, 1) + .1)        # Increase left side plot margins
plot(x = sim$p[sim$mu != 0], 
     y = rep(1, n.H1) + rnorm(n.H1, mean = 0, sd = .1),
     ylim = c(.5, 2), col = "red", xlab = "p value", ylab = "", yaxt = "n", xlim = c(0, 1))

points(x = sim$p[sim$mu == 0],
       y = rep(1.5, n.H0) + rnorm(n.H0, mean =0, sd = .1), col = "blue")

mtext("Null = TRUE", side = 2, at = 1.5, las = 1, line = 1)
mtext("Null = FALSE", side = 2, at = 1, las = 1, line = 1)

abline(v = .05, lty = 2)

# Add plot title
mtext(paste("p(Null is FALSE | p < .05) = ", round(with(subset(sim, p < .05), mean(mu != 0)), 2)), 
      side = 3, line = 1)
```
